---
title: "Diabetes"
author: "Manuel Tiburtini"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

## **About Dataset**

### **Context**

This dataset is originally from the National Institute of Diabetes and Digestive and Kidney Diseases. The objective is to predict based on diagnostic measurements whether a patient has diabetes.

### **Content**

Several constraints were placed on the selection of these instances from a larger database. In particular, all patients here are females at least 21 years old of Pima Indian heritage.

-   **Pregnancies**: Number of times pregnant 

-   **Glucose**: Plasma glucose concentration a 2 hours in an oral glucose tolerance test 

-   **BloodPressure**: Diastolic blood pressure (mm Hg) 

-   **SkinThickness**: Triceps skin fold thickness (mm) 

-   **Insulin**: 2-Hour serum insulin (mu U/ml) 

-   **BMI**: Body mass index (weight in kg/(height in m)\^2) 

-   **DiabetesPedigreeFunction**: Diabetes pedigree function that calculates diabetes likelihood depending on the subject's age and his/her diabetic family history.

-   **Age**: Age (years) 

-   **Outcome**: Class variable (0 or 1), where 0 is no diabetes and 1 is diabetes

#### **Sources:**

(a) Original owners: National Institute of Diabetes and Digestive and\
    Kidney Diseases\
(b) Donor of database: Vincent Sigillito (vgs\@aplcen.apl.jhu.edu)\
    Research Center, RMI Group Leader\
    Applied Physics Laboratory\
    The Johns Hopkins University\
    Johns Hopkins Road\
    Laurel, MD 20707\

<!-- -->

(301) 953-6231\

<!-- -->

(c) Date received: 9 May 1990

## Let's get started

```{r Loading packages}
start <- Sys.time()
library(conflicted) #resolving conflicting functions
library(tidyverse) #for tidy data manipulation
conflict_prefer("precision", "precision")
```

### Importing data

```{r importing the data}

Diabetes <- read.csv("~/Desktop/Material TADAB Workshop/Datasets/Diabetes/diabetes.csv")



```

### Data inspection

```{r converting the outcome to a factor with 2 levels}

glimpse(Diabetes)

Diabetes <- Diabetes %>% 
  mutate(diabetes = ifelse(Outcome == 1, "YES", "NO"), .keep="unused") %>% 
  mutate(diabetes = as.factor(diabetes))

```

```{r checking type of data}
library(visdat)

vis_dat(Diabetes)

#type of data
```

```{r checking missing data}
vis_miss(Diabetes)

#the data is all observed
```

### Data visualization

```{r Visualizing the data}
library(GGally) # paired plots 
library(ggthemes) #fancy themes

#visualizing paired data
ggpairs(Diabetes, columns = 1:9, ggplot2::aes(colour=diabetes, alpha=0.4), 
        lower=list(combo=wrap("facethist", binwidth=10)), progress = FALSE)+
  theme_minimal()+
  theme(panel.grid=element_blank())

# in the data are present zeros

```

```{r checking the distribution of zeros}

Diabetes %>% 
  summarize_all(~ sum(. == 0)) %>% 
  pivot_longer(everything(), names_to = "feature", values_to = "zeros") %>% 
  arrange(desc(zeros)) %>% 
  filter(zeros>0) %>% 
  mutate(perc_zeros=zeros/nrow(Diabetes)) %>%
  ggplot(aes(x=reorder(feature, desc(perc_zeros)), y=perc_zeros, fill=feature))+
  geom_bar(stat="identity", color="black")+
  xlab("Features")+
  ylab("zeros")+
  scale_y_continuous(labels = function(x) paste0(x*100, "%"))+
  theme_minimal()

```

Zeros are common, different choices can be made here, for simplicity, we keep them as zeros.

### Data splitting 

```{r splitting data}
#load the metapackage
library(tidymodels)
require(rsample)
#data split
set.seed(123)
diabetes_split <- initial_split(Diabetes, strata = "diabetes")

#check
diabetes_split

#split
diabetes_train <- training(diabetes_split)
diabetes_test <- testing(diabetes_split)

```

### Data visualization: how much are variables correlated?

```{r visualizing correlations}
library(corrplot)
require(grDevices)

corr_cols <- colorRampPalette(c("#91CBD765", "#CA225E"))

#correlation plot
diabetes_train %>%
    select(-diabetes) %>%
    cor() %>%
    corrplot(col = corr_cols(200), tl.col = "black", method =
"ellipse")

#correlation mean 
diabetes_train %>%
    select(-diabetes) %>%
    cor() %>%
    data.frame() %>% 
    summarise(across(where(is.numeric), .fns = 
                     list(mean = mean,
                          stdev = sd))) %>% 
    pivot_longer(everything(), names_sep='_', names_to=c('variable', '.value'))

```

### Setting the Logistic regression engine

```{r Set the engine!}
require(parsnip)
glm_model <- logistic_reg() %>% 
  set_engine("glm") %>%
  set_mode("classification") 

#check the engine glm
glm_model

```

### Recipes : preprocessing data

#### When setting the recipe....

\
recipe(diabetes \~ Pregnancies + Glucose + BloodPressure + SkinThickness + Insulin + BMI + DiabetesPedigreeFunction + Age, data = diabetes_train)\
\
is the same as writing\
\
(diabetes \~ ., data = diabetes_train)

```{r Set up the receipe}

diabetes_recipe <- 
 recipe(diabetes ~  Pregnancies + Glucose + BloodPressure + SkinThickness + Insulin + BMI + DiabetesPedigreeFunction + Age, # Specify the model
    data = diabetes_train
  ) 
#the simplest recipe

# if we would like to transform the data with step_log, we can juice the recipe and check the result

#juice(diabetes_prep) %>% 
#  ggpairs(., columns = 1:9, ggplot2::aes(colour=diabetes, alpha=0.4), 
#        lower=list(combo=wrap("facethist", binwidth=10)), progress = FALSE)+
#  theme_minimal()+
#  theme(panel.grid=element_blank())
  
#check the step that will be done by the model 
diabetes_recipe
```

### Workflow

```{r Set up the workflow }

#prepare the workflow
diabetes_workflow <-
  workflow() %>%
  add_model(glm_model) %>%
  add_recipe(diabetes_recipe)


#check the workflow
diabetes_workflow
```

### Extract predictions from the training set

```{r Fit and Predict on Training Data}

#fit the model
diabetes_fit_train <- fit(diabetes_workflow, diabetes_train)

#make predictions
diabetes_train_pred <-  predict(diabetes_fit_train, 
          new_data = diabetes_train %>% select(-diabetes), type = "prob") %>%
  bind_cols(
    truth = factor(diabetes_train$diabetes),
    .,
    predict(diabetes_fit_train, 
            new_data = diabetes_train %>% select(-diabetes), type = "class"))

diabetes_train_pred
```

#### Metrics of the training set

```{r Metrics on Training Data}


require(yardstick)
#confusion matrix
diabetes_train_pred %>% conf_mat(truth = truth, estimate = .pred_class)

#accuracy
diabetes_train_pred %>% accuracy(., truth= truth, estimate =.pred_class)

#metric set
classification_metrics <- metric_set(precision,recall, accuracy, mcc, f_meas) 
classification_metrics(diabetes_train_pred, truth = truth, estimate = .pred_class)

```

### Fit on the test set

```{r Fit and Predict on Test Data}

#fit the model 
diabetes_fit_test <-  fit(diabetes_workflow, diabetes_test)

diabetes_fit_test

#make predictions
diabetes_test_pred <- 
 predict(diabetes_fit_test, 
          new_data = diabetes_test %>% select(-diabetes), type = "prob") %>%
  bind_cols(
    truth = factor(diabetes_test$diabetes),
    .,
    predict(diabetes_fit_test, 
            new_data = diabetes_test %>% select(-diabetes), type = "class"))

diabetes_test_pred
```

#### Metrics of the test set

```{r Metrics on Test Data}

require(yardstick)
#confusion matrix
diabetes_test_pred %>% conf_mat(truth = truth, estimate = .pred_class)

#accuracy
diabetes_test_pred %>% accuracy(., truth= truth, estimate =.pred_class)

diabetes_test_pred %>% roc_auc(truth = truth, .pred_NO)

#metric set
classification_metrics <- metric_set(accuracy, mcc, f_meas) 
classification_metrics(diabetes_test_pred, truth = truth, estimate = .pred_class)

```

### ROC analysis

```{r Comparing the trainin and testing with roc}

require(yardstick)
train_auc <- 
  diabetes_train_pred %>% 
  roc_curve(truth = truth, .pred_NO)  %>% 
  mutate(Dataset = "Training set")

#Yuden J point for testing dataset
diabetes_train_pred %>% 
j_index(truth, .pred_class)


testing_auc <- 
  diabetes_test_pred %>% 
  roc_curve(truth = truth, .pred_NO) %>% 
    mutate(Dataset = "Test set")

#Yuden J point for testing dataset
diabetes_test_pred %>% 
j_index(truth, .pred_class)

```

```{r Plotting the ROC curves}

bind_rows(train_auc, testing_auc) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity, col = Dataset)) + 
  geom_path(lwd = 1.5, alpha = 0.8) +
  geom_abline(lty = 3) + 
  coord_equal() + 
  scale_color_viridis_d(option = "plasma", end = .6)+
  theme_clean()

Sys.time() - start 
##END running time 10.47secs
```

## Conclusion 

Here we have shown how to train a logististic regression model for classification using tidymodels, split the data and test on the test set comparing the results with a ROC.
